---
title: "Data Science 1, Homework 2"
author: "Emil Hafeez (eh2928)"
date: "9/24/2020"
output: github_document
---
## Read in the libraries
```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
```

# Problem One

#Read in the Mr Trash Wheel data, and process it.
Here, we assign the data in the Mr. Trashwheel sheet of an Excel sheet to a variable. We specify the data in the sheet, as well as clean the variable names, drop the rows without dumpster-specific data (which appear to be subtotals in the original sheet), and then round a variable to the nearest integer and store that variable respectively as an integer. That's it! In 10 lines! 
```{r trashwheel df and cleaning}
trashwheel_df = 
    read_xlsx(
      path =  "./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
      sheet = "Mr. Trash Wheel",
      range = cell_cols("A:N")) %>% 
    janitor::clean_names() %>% 
    drop_na(dumpster) %>% 
	mutate(
		sports_balls = round(sports_balls),
		sports_balls = as.integer(sports_balls)
      )
```
#Implement a similar process for the 2017 and 2018 precipitation data sheets from the same file. 
We implement a similar cleaning process as above, though instead for two years of precipitation data. We also omit rows without precipitation data, and also add a variable for the year. We also combine the 2018 and 2018 datasets, and convert the month into a character variable.
```{r precipitation_2017}
precipitation_2017_df = 
    read_xlsx(
      path =  "./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
      sheet = "2017 Precipitation", 
		  skip = 1) %>% 
    janitor::clean_names() %>% 
    drop_na(month) %>% 
	mutate(year = 2017) %>% 
  relocate(year, before = month)
```

```{r precipitation_2018}
precipitation_2018_df = 
    read_xlsx(
      path =  "./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
      sheet = "2018 Precipitation", 
		  skip = 1) %>% 
    janitor::clean_names() %>% 
    drop_na(month) %>% 
	mutate(year = 2018) %>% 
  relocate(year, before = month)
```










